---
title: "sewanee_climate"
author: "Linnea Minich"
date: "2024-06-12"
output: html_document
---

# Introduction

### Questions

1. Average precipitation for each month in five year periods.
2. Effect of temperature on precipitation from 2010 to 2020
3. Minimum and maximum temperature over time for the entire period
4. How has snowfall changed over time for the entire period?
5. Average temperature over time.
6. What month has the most snowfall on average

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#load library
library(tidyverse)
setwd("/Users/linneaminich/Desktop/linnea_datalab/datalab_2024/sewanee_climate/")
#load data
weather <- read_csv("sewanee_weather.csv")
```

# Average precipitation for each month in five year periods.

```{r}
#create new column named year for years and month for months
weather<-weather %>%
  mutate(year=year(ymd(DATE)),
         month=month(ymd(DATE)),
         month_name=month.abb[month]) %>%
  mutate(month_name=fct_reorder(month_name,month))
#create names for year groups
weather <- weather %>% 
  mutate(yrblock = case_when(
    year >=2000 & year <= 2004 ~ "2000-2004",
    year >=2005 & year <= 2009 ~ "2005-2009",
    year >=2010 & year <= 2014 ~ "2010-2014",
    year >=2015 & year <= 2019 ~ "2015-2019",
    year >=2020 & year <= 2024 ~ "2020-2024",
  .default = "other"
  ))
```

```{r}
#create a new datasets with months, year blocks, and average precipitation
precip <- weather %>% 
  group_by(yrblock, month_name) %>% 
  summarize(meanprecip = mean(PRCP, na.rm = TRUE))
```

```{r}
#Graphs
ggplot(precip, aes(x = month_name, y=meanprecip))+
  geom_col(fill="lightblue")+
  facet_wrap(~yrblock, nrow = 5)
  labs(x = "Month", y = "Average Precipitation")
```

# Average precipication and average temperature

```{r}
temp <- weather %>% 
  group_by(yrblock, month_name) %>% 
  summarize(meantemp = mean(TMAX, na.rm = TRUE))
```

```{r}
ggplot(temp, aes(x = month_name, y=meantemp))+
  geom_col(fill="#FF7074")+
  facet_wrap(~yrblock, nrow = 5)
  labs(x = "Month", y = "Average Temperature")
```

```{r}
tempmin <- weather %>% 
  group_by(yrblock, month_name) %>% 
  summarize(meantemp = mean(TMIN, na.rm = TRUE))
```

```{r}
ggplot(tempmin, aes(x = month_name, y=meantemp))+
  geom_col(fill="pink")+
  facet_wrap(~yrblock, nrow = 5)
  labs(x = "Month", y = "Average Temperature")
```
